---
title: "R Tidyverse for Marketing Analytics"
subtitle: "An Interactive Textbook"
author: "University of Liverpool Management School"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    scrollable: true
    hash: true
    chalkboard:
      buttons: true
    multiplex: true
    footer: "R Tidyverse · Marketing Analytics"
execute:
  echo: true
  warning: false
  message: false
---

# What is Tidyverse?

## The Tidyverse Philosophy

> "A collection of R packages designed for data science, sharing an underlying design philosophy and grammar."

**Core idea:** Write code the way you *think* about data analysis.

```{r}
#| eval: false
library(tidyverse)   # loads 8 core packages at once
```

## The 8 Core Packages

| Package | Purpose | Marketing Use Case |
|---------|---------|-------------------|
| `dplyr` | Data manipulation | Filter customers, compute metrics |
| `ggplot2` | Visualisation | Campaign performance charts |
| `tidyr` | Reshape data | Wide ↔ long format for modelling |
| `readr` | Read data | Load CSV files fast |
| `purrr` | Functional programming | Batch process survey responses |
| `tibble` | Modern data frames | Better printing & subsetting |
| `stringr` | String manipulation | Clean product names, parse URLs |
| `forcats` | Factor handling | Reorder brand categories |

## Installation

```{r}
#| eval: false
install.packages("tidyverse")   # one command, everything included
```

That's it. Now let's learn by doing.

---

# The Pipe Operator `%>%`

## Thinking in Pipes

**Without pipes** — nested functions, read inside-out:

```{r}
#| eval: false
result <- round(mean(filter(data, group == "A")$revenue), 2)
```

**With pipes** — left-to-right, like a sentence:

```{r}
#| eval: false
result <- data |>
  filter(group == "A") |>
  pull(revenue) |>
  mean() |>
  round(2)
```

> Think of `|>` as **"and then..."**

## Your Turn {background-color="#f0f4f8"}

What does this pipeline do?

```{r}
#| eval: false
customers |>
  filter(churn == "Yes") |>
  group_by(contract_type) |>
  summarise(
    n = n(),
    avg_tenure = mean(tenure_months)
  ) |>
  arrange(desc(n))
```

:::{.callout-note}
**Answer:** It finds churned customers, counts them by contract type,
calculates average tenure for each group, then sorts by count (largest first).
This tells you which contract type loses the most customers.
:::

---

# Reading Data with `readr`

## Load Marketing Data

```{r}
library(readr)

# Simulated customer dataset
customers <- tibble(
  customer_id = 1001:1012,
  segment = c("Premium", "Standard", "Budget", "Premium", "Standard",
              "Budget", "Premium", "Standard", "Budget", "Premium",
              "Standard", "Budget"),
  revenue = c(5200, 3100, 1800, 4800, 2900, 1500, 6100, 3300, 1200, 5500, 2700, 1600),
  churn = c("No", "No", "Yes", "No", "Yes", "Yes", "No", "No", "Yes", "No", "No", "Yes"),
  tenure_months = c(36, 24, 6, 30, 12, 3, 48, 18, 2, 42, 15, 4)
)

customers
```

## Reading Real Files

```{r}
#| eval: false
# CSV — the most common format
sales <- read_csv("data/sales_2025.csv")

# TSV — tab-separated
survey <- read_tsv("data/survey_responses.tsv")

# Specifying column types manually
campaigns <- read_csv("data/campaigns.csv",
  col_types = cols(
    campaign_id = col_character(),
    spend = col_double(),
    impressions = col_integer(),
    date = col_date(format = "%Y-%m-%d")
  )
)
```

:::{.callout-tip}
Always use `read_csv()` (from readr), **not** `read.csv()` (base R).
It's faster, gives a progress bar, and handles types better.
:::

---

# `dplyr`: The 6 Verbs

## The Grammar of Data Manipulation

```
filter()    → pick rows    (which customers?)
select()    → pick columns (which variables?)
mutate()    → new columns  (compute metrics)
arrange()   → sort rows    (rank by performance)
summarise() → collapse     (aggregate metrics)
group_by()  → stratify     (per segment)
```

These 6 verbs cover **90% of data wrangling** in marketing analytics.

---

## `filter()` — Pick Rows

```{r}
# Premium customers only
customers |>
  filter(segment == "Premium")
```

## Multiple Conditions

```{r}
# Premium customers with high revenue who haven't churned
customers |>
  filter(
    segment == "Premium",
    revenue > 5000,
    churn == "No"
  )
```

```{r}
# Customers who churned OR have low tenure
customers |>
  filter(churn == "Yes" | tenure_months <= 3)
```

> **Marketing interpretation:** This identifies at-risk customers —
> those who already left or are very new. Target these for retention campaigns.

---

## `select()` — Pick Columns

```{r}
# Only the columns you need
customers |>
  select(customer_id, segment, revenue)
```

## Helper Functions

```{r}
#| eval: false
# Select by pattern
select(data, starts_with("campaign_"))
select(data, ends_with("_rate"))
select(data, contains("click"))

# Exclude columns
select(data, -customer_id, -timestamp)

# Rename while selecting
select(data, seg = segment, rev = revenue)
```

:::{.callout-note}
**Why select?** Working with fewer columns makes code clearer,
prints neater, and prevents accidental information leakage in modelling.
:::

---

## `mutate()` — Create New Columns

```{r}
customers |>
  mutate(
    revenue_per_month = revenue / tenure_months,
    is_high_value = revenue > 4000,
    risk_level = case_when(
      churn == "Yes" ~ "Lost",
      tenure_months <= 6 ~ "High Risk",
      revenue > 4000 ~ "Stable",
      TRUE ~ "Monitor"
    )
  )
```

> `case_when()` is your best friend for creating marketing segments —
> it's readable, flexible, and handles edge cases with `TRUE ~`.

## `mutate()` Across Columns

```{r}
#| eval: false
# Apply same transformation to multiple columns
campaign_data |>
  mutate(across(contains("_rate"), ~ .x * 100))  # decimals to percentages

# Or with a named list
campaign_data |>
  mutate(across(
    c(open_rate, click_rate, conversion_rate),
    list(pct = ~ .x * 100, label = ~ paste0(round(.x * 100, 1), "%"))
  ))
```

---

## `arrange()` — Sort Rows

```{r}
# Highest revenue first
customers |>
  arrange(desc(revenue))
```

```{r}
# Sort by segment, then by revenue within each segment
customers |>
  arrange(segment, desc(revenue))
```

---

## `summarise()` + `group_by()`

```{r}
# Revenue per segment
customers |>
  group_by(segment) |>
  summarise(
    n_customers = n(),
    total_revenue = sum(revenue),
    avg_revenue = mean(revenue),
    churn_rate = mean(churn == "Yes")
  ) |>
  arrange(desc(total_revenue))
```

> `mean(churn == "Yes")` is a neat trick — TRUE/FALSE becomes 1/0, so the mean
> gives you the **churn rate** directly. Very useful for binary marketing outcomes.

## Multiple Grouping Levels

```{r}
#| eval: false
# Revenue by segment AND churn status
customers |>
  group_by(segment, churn) |>
  summarise(
    n = n(),
    avg_revenue = mean(revenue),
    avg_tenure = mean(tenure_months),
    .groups = "drop"     # always drop grouping after summarise
  )
```

---

## Combining All Verbs

A typical marketing analysis pipeline:

```{r}
customers |>
  filter(churn == "No") |>                          # active customers only
  mutate(tenure_years = tenure_months / 12) |>      # compute new metric
  group_by(segment) |>                              # per segment
  summarise(                                        # aggregate
    n = n(),
    avg_revenue = mean(revenue),
    avg_tenure_yr = mean(tenure_years),
    clv_estimate = avg_revenue * avg_tenure_yr      # rough CLV
  ) |>
  arrange(desc(clv_estimate))                       # rank
```

> **Customer Lifetime Value (CLV)** is one of the most important metrics in marketing.
> Here we use a simple formula: `revenue × tenure`. In practice, use discount rates and churn probabilities.

---

# `tidyr`: Reshape Data

## Why Reshape?

Marketing data comes in two shapes:

```{r}
# WIDE — one row per entity, metrics in columns (common in reports)
wide_data <- tibble(
  brand = c("Nike", "Adidas", "Puma"),
  q1_spend = c(120, 95, 40),
  q2_spend = c(135, 88, 52),
  q3_spend = c(140, 102, 48)
)

wide_data
```

## Wide → Long with `pivot_longer()`

```{r}
long_data <- wide_data |>
  pivot_longer(
    cols = starts_with("q"),
    names_to = "quarter",
    values_to = "spend",
    names_prefix = "q",
    names_transform = list(quarter = as.integer)
  )

long_data
```

> **ggplot2 requires long format.** If your chart looks wrong, pivot longer first.

## Long → Wide with `pivot_wider()`

```{r}
# Back to wide
long_data |>
  pivot_wider(
    names_from = quarter,
    values_from = spend,
    names_prefix = "q"
  )
```

## Handling Missing Data

```{r}
#| eval: false
# Fill missing values
data |>
  fill(campaign_name, .direction = "down")   # forward-fill

# Drop rows with missing values
data |>
  drop_na(revenue)                           # only where revenue is NA

# Replace NAs
data |>
  replace_na(list(spend = 0, channel = "Unknown"))

# Explicit NAs in pivoting
long_data |>
  pivot_wider(
    names_from = quarter,
    values_from = spend,
    values_fill = 0          # replace NA with 0
  )
```

---

# `ggplot2`: Visualise

## The Grammar of Graphics

Every ggplot has **three layers minimum**:

```{r}
#| eval: false
ggplot(data = <data>) +         # 1. What data?
  aes(x = <var>, y = <var>) +   # 2. What variables?
  geom_<type>()                  # 3. What geometry?
```

> Think: **data → aesthetics → geometry**. Same philosophy as the ggplot you learned before, now in R.

## Bar Chart — Revenue by Segment

```{r}
#| fig-width: 8
#| fig-height: 4
library(ggplot2)

customers |>
  group_by(segment) |>
  summarise(total_revenue = sum(revenue)) |>
  ggplot(aes(x = reorder(segment, total_revenue), y = total_revenue, fill = segment)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::dollar(total_revenue)), hjust = 1.2, colour = "white", size = 4) +
  coord_flip() +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Total Revenue by Customer Segment",
    subtitle = "Premium segment generates the most revenue",
    x = NULL,
    y = "Total Revenue (£)"
  ) +
  theme_minimal(base_size = 14)
```

## Scatter Plot — Revenue vs Tenure

```{r}
#| fig-width: 8
#| fig-height: 5
customers |>
  ggplot(aes(x = tenure_months, y = revenue, colour = churn)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  scale_colour_manual(values = c("No" = "#2c7bb6", "Yes" = "#d7191c")) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "£")) +
  labs(
    title = "Revenue vs Customer Tenure",
    subtitle = "Churned customers cluster at low tenure",
    x = "Tenure (months)",
    y = "Revenue (£)",
    colour = "Churned?"
  ) +
  theme_minimal(base_size = 14)
```

> **Marketing insight:** Churned customers (red) are concentrated in the
> lower-left — low tenure, low revenue. This suggests an **early retention problem**,
> not a product quality issue.

## Faceting — Small Multiples

```{r}
#| fig-width: 9
#| fig-height: 4
customers |>
  ggplot(aes(x = tenure_months, y = revenue)) +
  geom_point(size = 3, aes(colour = churn)) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ segment) +
  scale_colour_manual(values = c("No" = "#2c7bb6", "Yes" = "#d7191c")) +
  labs(
    title = "Revenue vs Tenure by Segment",
    x = "Tenure (months)",
    y = "Revenue (£)"
  ) +
  theme_minimal(base_size = 13)
```

---

# `stringr` & `forcats`

## Clean Text with `stringr`

```{r}
#| eval: false
library(stringr)

# Common marketing data cleaning
products |>
  mutate(
    name_clean = str_to_title(product_name),     # "nike AIR max" → "Nike Air Max"
    brand = str_extract(name_clean, "^\\w+"),    # first word as brand
    price_num = str_extract(price, "\\d+") |>    # "£120" → "120"
              as.numeric(),
    has_discount = str_detect(promo, "sale|off") # TRUE/FALSE
  )
```

## Reorder Categories with `forcats`

```{r}
#| fig-width: 8
#| fig-height: 4

# fct_reorder: order bars by value (not alphabetically)
customers |>
  mutate(segment = fct_reorder(segment, revenue, .fun = mean)) |>
  ggplot(aes(x = segment, y = revenue, fill = segment)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Revenue Distribution by Segment",
    subtitle = "Segments ordered by median revenue",
    x = NULL,
    y = "Revenue (£)"
  ) +
  theme_minimal(base_size = 14)
```

---

# Real-World Workflow

## Campaign Performance Analysis

```{r}
# A realistic marketing dataset
campaigns <- tibble(
  campaign = c("Summer Sale", "Summer Sale", "Black Friday", "Black Friday",
               "Loyalty Email", "Loyalty Email", "Social Ads", "Social Ads"),
  channel = c("Email", "Social", "Email", "Social", "Email", "Social", "Email", "Social"),
  impressions = c(50000, 120000, 80000, 200000, 30000, 45000, 40000, 95000),
  clicks = c(2500, 3600, 5600, 8000, 1800, 1350, 1600, 4750),
  conversions = c(125, 72, 392, 160, 198, 45, 64, 95),
  spend = c(500, 1800, 800, 3000, 200, 675, 400, 1425)
)
```

## Step 1: Compute Key Metrics

```{r}
campaign_metrics <- campaigns |>
  mutate(
    ctr = clicks / impressions,              # click-through rate
    cvr = conversions / clicks,              # conversion rate
    cpc = spend / clicks,                    # cost per click
    cpa = spend / conversions,               # cost per acquisition
    roas = (conversions * 50) / spend        # return on ad spend (£50 avg order)
  ) |>
  arrange(desc(roas))

campaign_metrics |>
  select(campaign, channel, ctr, cvr, cpa, roas) |>
  print(width = Inf)
```

## Step 2: Visualise Performance

```{r}
#| fig-width: 9
#| fig-height: 5
campaign_metrics |>
  mutate(campaign = fct_reorder(campaign, roas)) |>
  ggplot(aes(x = roas, y = campaign, fill = channel)) +
  geom_col(alpha = 0.85) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "red", linewidth = 0.7) +
  annotate("text", x = 1.1, y = 3.5, label = "Break-even", colour = "red", size = 3.5) +
  scale_x_continuous(labels = scales::dollar_format(prefix = "£")) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Campaign ROAS by Channel",
    subtitle = "Dashed line = break-even; everything to the right is profitable",
    x = "Return on Ad Spend (£)",
    y = NULL,
    fill = "Channel"
  ) +
  theme_minimal(base_size = 14)
```

## Step 3: Summarise & Recommend

```{r}
campaigns |>
  group_by(channel) |>
  summarise(
    total_spend = sum(spend),
    total_conversions = sum(conversions),
    overall_cpa = total_spend / total_conversions,
    overall_roas = (total_conversions * 50) / total_spend
  ) |>
  mutate(recommendation = case_when(
    overall_roas > 2 ~ "Scale up",
    overall_roas > 1 ~ "Optimise",
    TRUE ~ "Pause & review"
  )) |>
  arrange(desc(overall_roas))
```

> **Marketing decision:** Email outperforms Social on ROAS despite lower spend.
> Recommendation: **reallocate 20% of Social budget to Email** and test higher send volumes.

---

# Exercises

## Exercise 1: Customer Segmentation

Using the `customers` dataset:

1. Create a new column `tenure_group` with buckets: `<6mo`, `6-12mo`, `12-24mo`, `24+mo`
2. Calculate churn rate by `segment` and `tenure_group`
3. Visualise with a grouped bar chart
4. Which segment × tenure combination has the highest churn risk?

:::{.callout hint collapse="true"}
## Hint

```{r}
#| eval: false
customers |>
  mutate(tenure_group = case_when(
    tenure_months < 6 ~ "<6mo",
    tenure_months < 12 ~ "6-12mo",
    tenure_months < 24 ~ "12-24mo",
    TRUE ~ "24+mo"
  )) |>
  group_by(segment, tenure_group) |>
  summarise(churn_rate = mean(churn == "Yes"), n = n(), .groups = "drop") |>
  ggplot(aes(x = tenure_group, y = churn_rate, fill = segment)) +
  geom_col(position = "dodge") +
  labs(title = "Churn Rate by Segment and Tenure")
```
:::

## Exercise 2: Campaign Analysis

Using the `campaigns` dataset:

1. Which campaign has the **best CPA**? The **worst**?
2. Is Email or Social more cost-effective overall?
3. Create a scatter plot of `spend` vs `conversions`, sized by `ROAS`. What pattern do you see?

## Exercise 3: Build a Dashboard Table

Create a summary table that a marketing manager could use weekly:

```{r}
#| eval: false
# Your pipeline here:
# - Filter to a specific time period
# - Group by campaign and channel
# - Calculate: spend, impressions, clicks, conversions, CTR, CVR, CPA, ROAS
# - Add a "status" column (On Track / At Risk / Underperforming)
# - Arrange by ROAS descending
# - Use gt() or kable() for a formatted table
```

---

# Cheat Sheet

## The One-Page Reference

```{r}
#| eval: false
# THE TIDYVERSE WORKFLOW

# 1. LOAD
library(tidyverse)

# 2. READ
data <- read_csv("file.csv")

# 3. TRANSFORM (dplyr + tidyr)
result <- data |>
  filter(condition) |>              # keep rows
  select(columns) |>                # keep columns
  mutate(new_col = f(existing)) |>  # create columns
  pivot_longer(cols, names_to, values_to) |>  # reshape
  group_by(category) |>             # stratify
  summarise(metric = fun(value)) |> # aggregate
  arrange(desc(metric))             # sort

# 4. VISUALISE (ggplot2)
result |>
  ggplot(aes(x = var1, y = var2)) +
  geom_<type>() +
  labs(title = "...", x = "...", y = "...")

# 5. COMMUNICATE (interpret in marketing terms!)
```

---

# Key Takeaways

## What You Learned Today

1. **`|>` (pipe)** — chain operations left-to-right, like a sentence
2. **6 dplyr verbs** — filter, select, mutate, arrange, summarise, group_by
3. **`pivot_longer/wider`** — reshape data for analysis and visualisation
4. **ggplot2 grammar** — data + aesthetics + geometry + facets + theme
5. **Always interpret** — every number should answer a marketing question

## Common Mistakes to Avoid

| Mistake | Fix |
|---------|-----|
| Forgetting `group_by()` before `summarise()` | Always group first when you want per-category stats |
| Using `=` instead of `==` in `filter()` | `==` tests equality; `=` is for assignment |
| Plotting wide format data | Pivot longer first: ggplot needs tidy data |
| Ignoring NAs | Use `na.rm = TRUE` or `drop_na()` explicitly |
| Not interpreting results | Every metric → "so what?" for marketing |

## Next Steps

- **Practice:** Complete the exercises above
- **Read:** [R for Data Science (2e)](https://r4ds.hadley.nz/) — free online textbook
- **Explore:** `?dplyr`, `?ggplot2` in R console for built-in help
- **Next session:** Advanced ggplot2 — themes, annotations, and publication-quality figures

---

# Appendix: Setup

## R Environment

```{r}
#| eval: false
# Install (run once)
install.packages(c("tidyverse", "scales", "gt"))

# Load (run every session)
library(tidyverse)
library(scales)
library(gt)

# Recommended settings
options(digits = 3, scipen = 20)   # avoid scientific notation
theme_set(theme_minimal(base_size = 14))  # clean default theme
```
